%%-*-latex-*-

\section{Conclusion}

We presented a formal review design of the BER. On purpose, we
abstracted away many low-level details in our model in order to
understand, capture and formalise what are, according to us, the main
characteristics of the BER. Therefore the further step would be to
refine our model, by explicitly providing the coding and decoding
functions for the primitive types, by reckoning with the various
string types etc. Also we did not present evidences that the rewriting
from the BER domain to its \core subset conserves code equivalence, as
pointed out in figure~\ref{model2}: this was a matter of room. We
nevertheless think that our work dispels clouds of suspicion --- if
any --- about the soundness of \ASN and the BER. More precisely, we
mean that the composition of encoding and decoding yields a value
which is equivalent to the original. The aim of our formal review
design is to raise user's confidence on a solid ground and we doubt
whether twenty more pages of formul{\ae} would have been a stronger
argument for the casual reader. Indeed, making explicit as many as
possible assumptions and checking their consistence is inherently
reassuring. The mere fact that we had to understand the rationale of
the BER and put it into mathematical formal{\ae} really brought to the
fore a new understanding. Also the interest in choosing a system of
inference rules to define our relationships is that this formalism
closes the gap between specifications and algorithms. Besides, the
suggested use of OCaml as an implementation language is motivated
because, as a descendant of a logic meta-language, it is precisely
suited to implement algorithms specified by means of inference
rules. The way of deducing them consists mainly in providing a
deterministic and constructive refinement which is sound and complete
with respect to the initial specification. By constructive we mean for
instance to replace existential quantifiers, the symmetry rule etc. by
explicit procedures, and determinism means, in the context of this
work, having no backtracking implied (\emph{e.g.,} no overlapping
conclusions).


